---
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(fig.width=6, fig.asp = 0.618, collapse=TRUE) 
```

### In-class exercises
### Unit 5: Spatial

***

### Exercise 1.1

Add bathymetric contour lines to the colored bathymetry raster map of the Gulf of Maine. Draw the contours at 50m, 250m and 1000m depths. Are there any basins in the GOM that go as deep as 1000m?

```{r, eval=FALSE}
# plot contours
GOM_bath_map_contours = ggplot()+
  geom_raster(data = bath_m , aes(x = x, y = y, fill = depth_m)) + 
  scale_fill_gradientn(colors=c("black", "darkblue", "lightblue"), 
                       values = scales::rescale(c(-6000, -300, 0)), # rescale to make 2 different gradients
                       name="Depth (m)") +
  geom_contour(data = bath_m, aes(x=x, y=y, z=depth_m), breaks=c(-50), size=c(0.25), colour="black") + # add 100m contour
  geom_contour(data = bath_m, aes(x=x, y=y, z=depth_m), breaks=c(-250), size=c(0.5), colour="black") + # add 250m contour
  geom_contour(data = bath_m, aes(x=x, y=y, z=depth_m), breaks=c(-1000), size=c(0.75), colour="black") + # add 250m contour
  geom_polygon(data = world_map, aes(x = long, y = lat, group = group), fill = "black", color = NA) + # add coastline
  coord_fixed(1.3, xlim = lon_bounds, ylim = lat_bounds, expand=FALSE) + # Crop map edges
  ylab("Latitude") + xlab("Longitude") + theme_classic()

GOM_bath_map_contours # print to screen
```

***

### Exercise 1.2

Using the raster stack we created earlier, crop out Cape Cod Bay. Turn this into a data frame. How does the mean chlorophyll a concentration at depths between 0 and -50m compare to the mean chlorophyll concentration between -50 and -100m?

```{r, eval=FALSE}
CCB_lon = c(-70.8, -69.8)
CCB_lat = c(41.6, 42.2)

CCB_raster = crop(raster_stack, extent(c(CCB_lon, CCB_lat)))
CCB_df = data.frame( raster::rasterToPoints(CCB_raster))

ggplot(aes(x=chl_a), data=CCB_df) +
  geom_histogram()

# find mean chl_a between depths of 0 and 50m
CCB_df %>%
  filter(bath_m <=0, bath_m > -50) %>%
  summarize(chl = mean(chl_a, na.rm=TRUE))
# find mean chl_a between depths of 50 and 100m
CCB_df %>%
  filter(bath_m <= -50, bath_m > -100) %>%
  summarize(chl = mean(chl_a, na.rm=TRUE))

# plot raster data
CCB_map = ggplot()+
  geom_raster(data = CCB_df, aes(x = x, y = y, fill = bath_m)) + 
  geom_polygon(data = world_map, aes(x = long, y = lat, group = group), fill = "darkgrey", color = NA) + # add coastline; group keeps multipolygon coordinates separated into distinct groups
  coord_fixed(1.3, xlim = CCB_lon, ylim = CCB_lat, expand=FALSE) + # Crop map edges
  ylab("Lat") + xlab("Lon") + theme_bw() 
CCB_map
```

***


